---
title: "607-wk6-project2-USSenate"
author: "Tyler Frankenberg"
date: "3/13/2021"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DT)

```

## Introduction

We often hear about the advantage of incumbency in American leglislative elections.  This data from senate races 1974 - 2018 shows the amount spent and number of candidates per party, with additional variables including which challenger vs incumbent, and outcome including the margin of victory.

This data comes from the Brookings Intitution's "Vital Statistics on Congress" at  <https://www.brookings.edu/multi-chapter-report/vital-statistics-on-congress/>.


```{r import}

rawdata <- as_tibble(read.csv("https://raw.githubusercontent.com/curdferguson/607-wk6-project2-US-senate/main/vitalstats_ch3_full.csv", na = ""))

```

## Data Cleaning - Steps:

1. Remove "disclaimer" notes at the bottom
2. Remove Row 1 and make years the column headers
3. get "outcome" as its own variable
4. get "Incumbent/ Challenger" as its own variable
5. get "party" as its own variable
6. get "n" and "amount" as separate values for each observation

```{r clean1}

# Use slice to remove row 1 and rows 48-end
rawdata <- rawdata %>% slice(2:47)

# Make years the column headers, name col 1, and drop years row and unnamed columns
# Remove blank rows and trim whitespace
rawdata[[1, 1]] <- "Col1"
names(rawdata) <- rawdata[1,]
rawdata <- rawdata %>% select(!c(3,6,8,10,13))
rawdata <- rawdata %>% slice(3:46) %>% filter(`Col1` != "" | `2018` != "")
rawdata$Col1<- rawdata$Col1 %>% str_trim(side = c("both"))

```
Before we can comfortably pivot this data, we need to ensure that all of the multiple variables currently housed in column 1 are given their own column, and that redundant data (such as total columns) is removed...

```{r col1_valuevectors}

# set factor levels for outcomes and incumbent/ challenger
outcome_levels <- c("Incumbent w/ >= 60%", "Incumbent w/ < 60%", "Challenger")
inc_chal_levels <- c("Incumbents", "Challengers")

# fill missing values
rawdata <- rawdata %>% fill(Col1, .direction = c("down"))

# add an "outcome" column
rawdata <- rawdata %>% mutate(
  Outcome = c(1:39))

# populate the "outcome" column with the appropriate factor level
rawdata$Outcome <- rawdata$Outcome %>% lapply(function(i) {
  if (i %in% 1:13){
      rawdata$Outcome[[i]] = outcome_levels[[1]]
    } else if (i %in% 14:26){
      rawdata$Outcome[[i]] = outcome_levels[[2]]
    } else if (i %in% 27:39){
      rawdata$Outcome[[i]] = outcome_levels[[3]]
    }}) %>% 
  factor(levels = outcome_levels)

# remove rows housing only a value for "outcome"
rawdata <- rawdata %>% filter(!is.na(`2018`))

# add a column for "incumbent/ challenger" and populate
rawdata <- rawdata %>% mutate(
  `Inc/Chal` = ifelse(
    rawdata$`Col1` == "Incumbents", "Incumbents", ifelse(
    rawdata$`Col1` == "Challengers", "Challengers", NA)))

# change "incumbent/ challenger" to type "factor" and set the levels
rawdata$`Inc/Chal` <- rawdata$`Inc/Chal` %>% factor(levels = inc_chal_levels)

# fill the "incumbent/ challenger" column
rawdata <- rawdata %>% fill(`Inc/Chal`, .direction = c("down"))

# retain only those rows for which col1 value is "democrats" or "republicans" and rename the column to "Party"
rawdata <- rawdata %>% filter(`Col1` == "Democrats" | `Col1` == "Republicans")
rawdata <- rename(rawdata, "Party" = `Col1`)
  

```


Since we have two sets of values which we'll ultimately want in our long tibble ("amount spent" and "n" or "number of candidates"), we'll accomplish the pivot and our last cleaning task by separating the two types of values out into separate tables.  We'll then join them back together once both are clean and in long format.


```{r split_data}

# get the even numbered rows where the messy "n" values are stored
split_data_n <- rawdata %>% filter((row(rawdata) %% 2) == 0)

# remove all information from "n" except their integer value
split_data_n[2:24] <- split_data_n[2:24] %>% lapply(function(i){
  str_match(i, "(?<==)[:digit:]{1,2}") %>% .[,1] %>% as.integer()
  })

# get the values for "years"
years <- names(split_data_n[2:24])

# pivot the "n" dataframe to a long format
split_data_n <- split_data_n %>% pivot_longer(cols=years, names_to="Year", values_to="N")

# get the odd numbered rows where the messy dollar values are stored
split_data_dol <- rawdata %>% filter((row(rawdata) %% 2) == 1)

# remove all commas from the dollar values
split_data_dol[2:24] <- split_data_dol[2:24] %>% lapply(function(i){
  str_replace_all(i, ",", "") %>% as.integer()
  })

# pivot the "dollars" dataframe to a long format 
split_data_dol <- split_data_dol %>% pivot_longer(cols=years, names_to="Year", values_to = "Total Spent")

```

```{r join_data}
# join the split, long-format data tables back together 
clean_data <- inner_join(split_data_n, split_data_dol)

```
